---
output: github_document
---


<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r setup, echo = FALSE, message = FALSE}
library(SomaDataIO)
library(ggplot2)
library(dplyr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
coverage <- "https://img.shields.io/badge/coverage-99-success.svg?style=flat&logo=codecov"
```


# The R `SomaDataIO` Package from SomaLogic, Inc.

<!-- badges: start -->
![cover](`r coverage`)
[![License: MIT](https://img.shields.io/badge/License-MIT-blue.svg)](https://choosealicense.com/licenses/mit/)
<!-- badges: end -->


-------------


## Overview

This document accompanies the R package `SomaDataIO`, which loads the 
SomaLogic, Inc. proprietary data file called an `*.adat`.
The package provides auxiliary functions for extracting relevant information 
from the ADAT object once in the R environment.
Basic familiarity with the R environment is assumed, as is the ability to install
contributed packages from the Comprehensive R Archive Network (CRAN).


-------------------


## Installation

1. Download the file `SomaDataIO_0.0.1.tar.gz`.

2. Ensure that the following package dependencies are pre-installed;
from the `R >= 3.6.0` console:

    ```{r pkgs, eval = FALSE}
    install.packages(c("magrittr", "devtools", "readr", "purrr", "usethis", "tidyr",
                       "dplyr", "tibble", "cli", "crayon", "stringr"))
    ```

3. Install via one of these methods:
    i. From the R console:
        ```{r install, eval = FALSE}
        install.packages("SomaDataIO_0.0.1.tar.gz", repos = NULL, type = "source")
        ```
    i. From a command-line terminal (Linux/Mac):
    
        ```bash
        R --vanilla CMD INSTALL SomaDataIO_0.0.1.tar.gz
        ```

4. **NOTE:** The `Biobase` package is *suggested*, being required by only 
one function. `Biobase` must be installed separately from `Bioconductor`
by entering the following from the `R` console:

    ```{r, eval = FALSE}
    if (!requireNamespace("BiocManager", quietly = TRUE)) {
      install.packages("BiocManager")
    }
    BiocManager::install("Biobase")
    ```

    Information about Bioconductor can be found here: 
  [https://bioconductor.org/install/](https://bioconductor.org/install/)


5. Upon *successful* installation:
    i. To load the package:
        ```{r load_somadataio, eval = FALSE}
        library(SomaDataIO)
        ```
    i. For an index of available commands:
        ```{r help, eval = FALSE}
        library(help = SomaDataIO)
        ``` 

---------------------




## Main Features

* Loading data (Import)
    + Import a text file in the `*.adat` format into an `R` session
      as a `soma_adat` object.
* Manipulating data
    + Subset, reorder, and list various fields of a `soma_adat` object.
* Exporting data (Output)
    + Write out a `soma_adat` object as a `*.adat` text file.


----------------------


## Load an ADAT
```{r read_adat}
# Sample file name
f <- system.file("sample", "sample.adat",
                 package = "SomaDataIO", mustWork = TRUE)
my_adat <- read_adat(f)
is.soma_adat(my_adat)

# S3 print method -> tibble
my_adat

print(my_adat, show_header = TRUE)  # if simply wish to see Header info, no features
```


## Attributes contain feature information
```{r atts}
attributes(my_adat)$Col.Meta
```


## Manipulate a `soma_adat` Object
```{r features}
getFeatures(my_adat) %>% head(20)     # first 20 features
getFeatures(my_adat) %>% length()     # how many features
getFeatures(my_adat, n = TRUE)        # the `n` argument; no. features
```


## Write out a `soma_adat`
```{r write}
is.intact.attributes(my_adat)         # attributes MUST be intact to write

fout <- tempfile(fileext = ".adat")   # random file name
write_adat(my_adat, file = fout)
```



-----------------


# A Typical Analysis
## Load data
```{r load}
data <- read_adat(f)
dim(data)

train <- dplyr::filter(data, TimePoint == "Young")
table(train$SampleGroup)    # M vs F
```

## Subset Features starting with `MMP`
```{r mmps}
mmps <- my_adat %>% 
  getFeatureData() %>% 
  filter(grepl("^MMP", Target))

# feature data info
mmps  
```

## Generate list of `t-tests`
```{r t_tests}
t_tests <- mmps$AptName %>% 
  purrr::set_names(., .) %>% 
  purrr::map(~ {
    sprintf("%s ~ SampleGroup", .x) %>% 
      as.formula() %>% 
      t.test(data = data)
  })
```

## Re-order by `p-value`
```{r p_values}
idx <- t_tests %>% 
  purrr::map("p.value") %>% 
  stringr::str_order()
idx

t_tests %<>% magrittr::extract(idx)
t_tests %>% head() %>% purrr::map("p.value")
```

## Visualize with `ggplot2()`
```{r ggplot_boxes, fig.width = 9, fig.height = 8}
targets <- as.list(mmps$AptName) %>%
  purrr::set_names(mmps$TargetFullName)   # create SeqId -> target map
df <- data %>% 
  dplyr::select(SampleGroup, all_of(mmps$AptName)) %>% 
  dplyr::rename(!!! targets) %>% 
  tidyr::pivot_longer(-SampleGroup, names_to = "Feature", values_to = "RFU") %>% 
  tibble::as_tibble()
df

df %>% 
  ggplot(aes(x = SampleGroup, y = log10(RFU), colour = SampleGroup)) +
  geom_boxplot(alpha = 0.5) +
  facet_wrap(~Feature)
```



---------------------

## MIT LICENSE
* See [LICENSE](LICENSE.md)
* The MIT License: 
  - [https://choosealicense.com/licenses/mit/](https://choosealicense.com/licenses/mit/)
  - [https://tldrlegal.com/license/mit-license/](https://tldrlegal.com/license/mit-license)


---------------------


Created by [Rmarkdown](https://github.com/rstudio/rmarkdown)
(v`r utils::packageVersion("rmarkdown")`) and `r R.version$version.string`. 
