---
output: github_document
---


<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r setup, echo = FALSE, message = FALSE}
options(width = 100)
library(SomaDataIO)
library(magrittr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
set.seed(1)   # set seet for ggplot jitter below; Git doesn't trigger every time
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
coverage <- paste0("https://img.shields.io/badge/coverage-",
                   80, "-success.svg?style=flat&logo=codecov")
ver <- desc::desc_get_version(".")
ver <- paste0("https://img.shields.io/badge/Version-", ver,
              "-success.svg?style=flat&logo=github")
#knitr::knit_exit()
```


# The `SomaDataIO` R Package from SomaLogic, Inc.

<!-- badges: start -->
![GitHub version](`r ver`)
[![CRAN badge](https://img.shields.io/badge/CRAN-No-red.svg)](https://cran.r-project.org)
![cover](`r coverage`)
[![License: MIT](https://img.shields.io/badge/License-MIT-blue.svg)](https://choosealicense.com/licenses/mit/)
<!-- badges: end -->


-------------


## Overview

This document accompanies the R package `SomaDataIO`, which loads the 
SomaLogic, Inc. proprietary data file called an `*.adat`.
The package provides auxiliary functions for extracting relevant information 
from the ADAT object once in the R environment.
Basic familiarity with the R environment is assumed, as is the ability to install
contributed packages from the Comprehensive R Archive Network (CRAN).


-------------------


## Installation

The easiest way to install `SomaDataIO` is to install directly from GitHub:

```{r install, eval = FALSE}
devtools::install_github("SomaLogic/SomaDataIO")
```

Alternatively you may clone the repository and install manually:

```bash
git clone https://github.com/SomaLogic/SomaDataIO.git SomaDataIO
R CMD INSTALL --use-vanilla --resave-data SomaDataIO
```


#### Package Dependencies

The `SomaDataIO` package was intentionally developed to run slightly behind
the bleeding edge of The Comprehensive R Archive Network (`CRAN`).
This allows lead time to identify and fix bugs as well 
as simplifying software life-cycle.
This may change in the future, however for the time being, the dependencies
below represent the environment in which `SomaDataIO` was developed.
If you run into any unexpected behavior, please
ensure that the following package dependencies are pre-installed:

* `R (>= 4.0.0)`
* `dplyr (v1.0.6)`
* `assertthat (v0.2.1)`
* `lifecycle (v1.0.0)`
* `magrittr (v2.0.1)`
* `devtools (v2.4.1)`
* `readr (v1.4.0)`
* `rlang (v0.4.11)`
* `purrr (v0.3.4)`
* `usethis (v2.0.1)`
* `tidyr (v1.1.3)`
* `tidyselect (v1.1.1)`
* `tibble (v3.2.1)`
* `cli (v2.5.0)`
* `crayon (v1.4.1)`
* `stringr (v1.4.0)`


#### Biobase

The `Biobase` package is *suggested*, being required by only 
two functions, `pivotExpressionSet()` and `adat2eset()`. `Biobase` must be 
installed separately from `Bioconductor` by entering the 
following from the `R` console:

```{r biobase, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("Biobase", version = "3.13")
```

Information about Bioconductor can be found here: [https://bioconductor.org/install/](https://bioconductor.org/install/)


#### Loading

Upon *successful* installation, load the SomaDataIO as normal:
```{r load-somadataio, eval = FALSE}
library(SomaDataIO)
```

For an index of available commands:
```{r help, eval = FALSE}
library(help = SomaDataIO)
``` 


#### Internal Objects

The `SomaDataIO` package comes with 4 internal objects available
to users to run canned examples (or analyses). They can be accessed once
`SomaDataIO` has been attached via `library()`. They are:

* `example_data`
* `ex_features`
* `ex_feature_data`
* `ex_target_names`
* See `?SomaDataObjects`


---------------------



## Main Features (I/O)

* Loading data (Import)
    + Import a text file in the `*.adat` format into an `R` session
      as a `soma_adat` object.
* Wrangling data (manipulation)
    + Subset, reorder, and list various fields of a `soma_adat` object.
* Exporting data (Output)
    + Write out a `soma_adat` object as a `*.adat` text file.


----------------------


### Loading an ADAT
```{r read-adat}
# Sample file name
f <- system.file("example", "example_data.adat", package = "SomaDataIO", mustWork = TRUE)
my_adat <- read_adat(f)
is.soma_adat(my_adat)

# S3 print method forwards -> tibble
my_adat

print(my_adat, show_header = TRUE)  # if simply wish to see Header info, no features

# S3 summary method
# View Target and summary statistics
seqs <- tail(names(my_adat), 3)
summary(my_adat[, seqs])

# Summarize by Sex
my_adat[, seqs] %>%
  split(my_adat$Sex) %>%
  lapply(summary)
```


### Wrangling 
#### Attributes Contain File and Feature Information
```{r atts}
names(attributes(my_adat))

# The `Col.Meta` attribute contains 
# target annotation information
attributes(my_adat)$Col.Meta
```

#### Feature Data
The `getFeatureData()` function creates a lookup table that links the feature
names in the `soma_adat` object to the annotation data in `Col.Meta` via the
common key, `AptName`, in column 1:

```{r annotations}
getFeatureData(my_adat)
```


#### Features (`seq.xxxx.xx`)
```{r feats}
getFeatures(my_adat) %>% head(20)     # first 20 features; see AptName above
getFeatures(my_adat) %>% length()     # how many features
getFeatures(my_adat, n = TRUE)        # the `n` argument; no. features
```

#### Clinical Data
```{r meta}
getMeta(my_adat)             # clinical meta data for each sample
getMeta(my_adat, n = TRUE)   # also an `n` argument
```

#### Math Generics
You may perform basic mathematical transformations on the feature data *only*
with special `soma_adat` S3 methods (see `?MathGenerics`):
```{r math-generics}
head(my_adat$seq.2429.27)

logData <- log10(my_adat)    # a typical log10() transform
head(logData$seq.2429.27)

roundData <- round(my_adat)
head(roundData$seq.2429.27)

sqData <- sqrt(my_adat)
head(sqData$seq.2429.27)
```

#### Full Complement of `dplyr` S3 Methods
The `soma_adat` also comes with numerous class specific methods to the most
popular `dplyr` generics that make working with `soma_adat` objects simpler
for those familiar with this standard toolkit:

```{r dplyr}
dim(my_adat)
males <- dplyr::filter(my_adat, Sex == "M")
dim(males)

males %>% 
  dplyr::select(SampleType, SampleMatrix, starts_with("NormScale"))
```

#### Available S3 Methods `soma_adat`
```{r methods}
# see full complement of `soma_adat` methods
methods(class = "soma_adat")
```

### Writing a `soma_adat`
```{r write}
is.intact.attributes(my_adat)     # attributes MUST be intact to write to file

write_adat(my_adat, file = tempfile("my-adat-", fileext = ".adat"))
```



-----------------


# Typical Analyses

Although it is beyond the scope of the `SomaDataIO` package, below are 3 
sample analyses that typical users/clients would perform on SomaLogic data.
They are not intended to be a definitive guide in statistical 
analysis and existing packages do exist in the `R` universe that perform parts
or extensions of these techniques. Many variations of the workflows below 
exist, however the framework highlights how one could perform standard
*preliminary* analyses on SomaLogic data for:

* Two-group differential expression (*t*-test)
* Binary classification (logistic regression)
* Linear regression

#### Data Preparation
```{r data-prep}
# `example_data` comes with SomaDataIO
dim(example_data)
table(example_data$SampleType)

is_seq <- function(.x) grepl("^seq\\.[0-9]{4}", .x) # regex for features

# Prepare data set for analysis
cleanData <- example_data %>%
  filter(SampleType == "Sample") %>%                # rm control samples
  tidyr::drop_na(Sex) %>%                           # rm NAs if present
  log10() %>%                                       # log10-transform (Math Generic)
  mutate(Group = as.numeric(factor(Sex)) - 1) %>%   # map Sex -> 0/1
  mutate_if(is_seq(names(.)), ~ {                   # mutate features only
    .x %>% subtract(mean(.)) %>% divide_by(sd(.))   # center & scale features
  })
table(cleanData$Sex)
table(cleanData$Group)    # F = 0; M = 1
```

## Compare Two Groups (M/F) via t-test
#### Get annotations via `getFeatureData()`:

```{r get-anno}
t_tests <- getFeatureData(cleanData) %>% 
  select(AptName, SeqId, Target = TargetFullName, EntrezGeneSymbol, UniProt)

# Feature data info:
#   Subset via dplyr::filter(t_tests, ...) here to 
#   restrict analysis to only certain analytes
t_tests
```

#### Calculate `t-tests`
Use a "list columns" approach via nested tibble object using `dplyr`, `purrr`,
and `stats::t.test()`

```{r t-tests}
t_tests <- t_tests %>% 
  dplyr::mutate(
    formula = map(AptName, ~ as.formula(paste(.x, "~ Sex"))), # create formula
    t_test  = purrr::map(formula, ~ stats::t.test(.x, data = cleanData)),  # fit t-tests
    t_stat  = purrr::map_dbl(t_test, "statistic"),            # pull out t-statistic
    p.value = purrr::map_dbl(t_test, "p.value"),              # pull out p-values
    fdr     = stats::p.adjust(p.value, method = "BH")         # FDR for multiple testing
  ) %>%
  dplyr::arrange(p.value) %>%                   # re-order by `p-value`
  dplyr::mutate(rank = dplyr::row_number())     # add numeric ranks

# View analysis tibble
t_tests
```

#### Visualize with `ggplot2()`
Create a plotting tibble in the "long" format for `ggplot2`:

```{r ggplot-data}
target_map <- head(t_tests, 12) %>%     # mapping table
  select(AptName, Target)               # SeqId -> Target

plot_tbl <- example_data %>%
  filter(SampleType == "Sample") %>%          # rm control samples
  drop_na(Sex) %>%                            # rm NAs if present
  log10() %>%                                 # log10-transform for plotting
  select(Sex, target_map$AptName) %>%         # top 12 analytes
  pivot_longer(cols = -Sex, names_to = "AptName", values_to = "RFU") %>% 
  dplyr::left_join(target_map) %>%
  # order factor levels by 't_tests' rank to order plots below
  mutate(Target = factor(Target, levels = target_map$Target))

plot_tbl
```

```{r ggplot-boxes, fig.width = 10, fig.height = 10}
plot_tbl %>%
  ggplot(aes(x = Sex, y = RFU, fill = Sex)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  scale_fill_manual(values = c("#24135F", "#00A499")) +
  geom_jitter(shape = 16, width = 0.1, alpha = 0.5) +
  facet_wrap(~ Target) +
  ggtitle("Boxplots of Top Analytes by t-test") +
  labs(y = "log10(RFU)") +
  theme(plot.title = element_text(size = 21, face = "bold"),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.position = "top"
  )
```


## Logistic Regression
#### Predict Sex

```{r logreg-tbl}
set.seed(3)                  # seed resulting in 50/50 class balance
idx   <- sample(1:nrow(cleanData), size = nrow(cleanData) - 50)  # hold-out
train <- cleanData[idx, ]
test  <- cleanData[-idx, ]

# assert no overlap
assertthat::assert_that(
  identical(intersect(rownames(train), rownames(test)), character(0))
)

LR_tbl <- getFeatureData(train) %>%
  select(AptName, SeqId, Target = TargetFullName, EntrezGeneSymbol, UniProt) %>%
  mutate(
    formula  = map(AptName, ~ as.formula(paste("Group ~", .x))),  # create formula
    model    = map(formula, ~ stats::glm(.x, data = train, family = "binomial", model = FALSE)),  # fit glm()
    beta_hat = map_dbl(model, ~ coef(.x)[2L]),      # pull out coef Beta
    p.value  = map2_dbl(model, AptName, ~ {
      summary(.x)$coefficients[.y, "Pr(>|z|)"] }),  # pull out p-values
    fdr      = p.adjust(p.value, method = "BH")     # FDR correction multiple testing
  ) %>%
  arrange(p.value) %>%            # re-order by `p-value`
  mutate(rank = row_number())     # add numeric ranks

LR_tbl
```


#### Fit Model | Calculate Performance
Next, select features for the model fit. We have a good idea of reasonable `Sex`
markers from prior knowledge (`CGA*`), and fortunately many of these are highly
ranked in `LR_tbl`. Below we fit a 4-marker logistic regression model from 
cherry-picked gender-related features:

```{r fit-logreg}
# AptName is index key between `LR_tbl` and `train`
feats <- LR_tbl$AptName[c(1, 3, 5, 7)]
form  <- as.formula(paste("Group ~", paste(feats, collapse = "+")))
fit   <- glm(form, data = train, family = "binomial", model = FALSE)
pred  <- tibble(
  true_class = test$Sex,                                         # orig class label
  pred       = predict(fit, newdata = test, type = "response"),  # prob. 'Male'
  pred_class = ifelse(pred < 0.5, "F", "M"),                     # class label
)
conf <- table(pred$true_class, pred$pred_class, dnn = list("Actual", "Predicted"))
tp   <- conf[2, 2]
tn   <- conf[1, 1]
fp   <- conf[1, 2]
fn   <- conf[2, 1]

# Confusion matrix
conf

# Classification metrics
tibble(Sensitivity = tp / (tp + fn),
       Specificity = tn / (tn + fp),
       Accuracy    = (tp + tn) / sum(conf),
       PPV         = tp / (tp + fp),
       NPV         = tn / (tn + fn)
)
```


## Linear Regression
We use the same `cleanData`, `train`, and `test` data objects from the
logistic regression analysis above.

#### Predict Age
```{r linreg-tbl}
LinR_tbl <- getFeatureData(train) %>%               # `train` from above
  select(AptName, SeqId, Target = TargetFullName, EntrezGeneSymbol, UniProt) %>%
  mutate(
    formula = map(AptName, ~ as.formula(paste("Age ~", .x, collapse = " + "))),
    model   = map(formula, ~ lm(.x, data = train, model = FALSE)),  # fit linear models
    slope   = map_dbl(model, ~ coef(.x)[2L]),       # pull out B_1
    p.value = map2_dbl(model, AptName, ~ {
      summary(.x)$coefficients[.y, "Pr(>|t|)"] }),  # pull out p-values
    fdr     = p.adjust(p.value, method = "BH")      # FDR for multiple testing
  ) %>%
  arrange(p.value) %>%           # re-order by `p-value`
  mutate(rank = row_number())    # add numeric ranks

LinR_tbl
```


#### Fit Model | Calculate Performance
Fit an 8-marker model with the top 8 features from `LinR_tbl`:

```{r linreg-fit}
feats <- head(LinR_tbl$AptName, 8)
form  <- as.formula(paste("Age ~", paste(feats, collapse = "+")))
fit   <- lm(form, data = train, model = FALSE)
n     <- nrow(test)
p     <- length(feats)

# Results
res   <- tibble(
  true_age   = test$Age,
  pred_age   = predict(fit, newdata = test),
  pred_error = pred_age - true_age
)

# Lin's Concordance Correl. Coef.
# Accounts for location + scale shifts
linCCC <- function(x, y) {
  stopifnot(length(x) == length(y))
  a <- 2 * cor(x, y) * sd(x) * sd(y)
  b <- var(x) + var(y) + (mean(x) - mean(y))^2
  a / b
}

# Regression metrics
tibble(
  rss  = sum(res$pred_error^2),                 # residual sum of squares
  tss  = sum((test$Age - mean(test$Age))^2),    # total sum of squares
  rsq  = 1 - (rss / tss),                       # R-squared
  rsqadj = max(0, 1 - (1 - rsq) * (n - 1) / (n - p - 1)), # Adjusted R-squared
  R2   = stats::cor(res$true_age, res$pred_age)^2,        # R-squared Pearson approx.
  MAE  = mean(abs(res$pred_error)),             # Mean Absolute Error
  RMSE = sqrt(mean(res$pred_error^2)),          # Root Mean Squared Error
  CCC  = linCCC(res$true_age, res$pred_age)     # Lin's CCC
)
```

#### Visualize Concordance
```{r linreg-plot, fig.width = 7, fig.height = 7}
lims <- range(res$true_age, res$pred_age)
res %>%
  ggplot(aes(x = true_age, y = pred_age)) +
  geom_point(colour = "#24135F", alpha = 0.5, size = 4) +
  expand_limits(x = lims, y = lims) +                # make square
  geom_abline(slope = 1, colour = "black") +         # add unit line
  geom_rug(colour = "#286d9b", size = 0.2) +
  labs(y = "Predicted Age", x = "Actual Age") +
  ggtitle("Concordance in Predicted vs. Actual Age") +
  theme(plot.title = element_text(size = 21, face = "bold"),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14))
```



---------------------

## MIT LICENSE
* See [LICENSE](LICENSE.md)
* The MIT License: 
  - [https://choosealicense.com/licenses/mit/](https://choosealicense.com/licenses/mit/)
  - [https://tldrlegal.com/license/mit-license/](https://tldrlegal.com/license/mit-license)


---------------------


Created by [Rmarkdown](https://github.com/rstudio/rmarkdown)
(v`r utils::packageVersion("rmarkdown")`) and `r R.version$version.string`. 
