---
title: "Command Line Merge Tool"
author: "Stu Field, SomaLogic Operating Co., Inc."
description: >
  A convenient CLI merge tool to add new clinical data
  to 'SomaScan' data.
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Command Line Merge Tool}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(SomaDataIO)
library(withr)
Sys.setlocale("LC_COLLATE", "en_US.UTF-8")
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```


# Overview

Occasionally, additional clinical data is obtained _after_ samples have been
submitted to SomaLogic, Inc. or even after 'SomaScan' results have
been delivered. 

This requires the new clinical, i.e. non-proteomic, data to be merged
with the 'SomaScan' data into a "new" ADAT prior to analysis.
For this purpose, a command-line interface ("CLI") tool has been included
with [SomaDataIO](https://CRAN.R-project.org/package=SomaDataIO)
in the `cli/merge/` directory, which allows one to
generate an updated `*.adat` file via the command-line without
having to open an integrated development environment ("IDE"), e.g. `RStudio`.


----------------


## Setup

The clinical merge tool is an `R script` that comes with an installation
of [SomaDataIO](https://CRAN.R-project.org/package=SomaDataIO):

```{r merge-script}
dir(system.file("cli", "merge", package = "SomaDataIO"))

f <- system.file("cli", "merge", "merge_clin.R", package = "SomaDataIO")
f
```

First create a temporary "analysis" directory:

```{r create-dir}
analysis_dir <- tempfile(pattern = "somascan-")
# create directory
dir.create(analysis_dir)

# sanity check
dir.exists(analysis_dir)

# copy merge tool into analysis directory
file.copy(f, to = analysis_dir)
```

Let's create some dummy 'SomaScan' data derived from the `example_data` object
from [SomaDataIO](https://CRAN.R-project.org/package=SomaDataIO)
First we reduce its size to 9 samples and 5 proteomic features each, and
then write to text with `write_adat()`.
This will be the "new" starting point for the clinical
data merge and represents where customers would begin their analysis.

```{r save-data}
feats <- withr::with_seed(3, sample(getAnalytes(example_data), 5))
sub_adat <- dplyr::select(example_data, PlateId, SlideId, Subarray,
                          SampleId, Age, all_of(feats)) |> head(9L)
withr::with_dir(analysis_dir,
  write_adat(sub_adat, file = "ex-data-9.adat")
)
```

Now we create random clinical data with a common key (this is typically
the `SampleId` identifier but it could be any common key).

```{r create-clin-1}
df <- data.frame(SampleId = as.character(seq(1, 9, by = 2)),  # common key
                 group    = c("a", "b", "a", "b", "a"),
                 newvar   = withr::with_seed(1, rnorm(5)))
df

# write clinical data to file
withr::with_dir(analysis_dir,
  write.csv(df, file = "clin-data.csv", row.names = FALSE)
)
```


At this point we have 3 files in a temporary analysis directory:

```{r ls1}
dir(analysis_dir)
```

1. `clin-data.csv`:
    + new data containing 3 columns:
    + a common key: `SampleId`
    + a new variable with grouping information: `group`
    + a new variable with a continuous variable: `newvar`
1. `ex-data-9.adat`:
    + ADAT with 9 samples containing 5 'SomaScan' proteomic
      features and 5 pre-existing variables we would like to merge into
    + `PlateId`, `SlideId`, `Subarray`, `SampleId`, and `Age`
    + __note:__  `PlateId`, `SlideId`, and `Subarray` are key fields common
      to _almost all_ ADATs; removing them could yield unintended results
    + the common key `SampleId` is required
1. `merge_clin.R` the merge script engine itself 


## Merging Clinical Data

The clinical data merge tool is simple to use at most common command line
terminals (`BASH`, `ZSH`, etc.). You must have `R` installed
(and available) with [SomaDataIO](https://CRAN.R-project.org/package=SomaDataIO)
and its dependencies installed.

### Arguments

The merge script takes 4 ordered arguments:

1. path to the original ADAT (`*.adat`) file
1. path to clinical data (`*.csv`) file
1. common key variable name (e.g. `SampleId`)
1. output file name (`*.adat`) for new ADAT



---------------


### Standard Syntax

The primary syntax is for when the common key in __both__ files,
(ADAT and CSV), has the _same_ variable name:


```bash
# change directory to the analysis path
cd `r analysis_dir`

# run the Rscript:
# - we recommend using the --vanilla flag
Rscript --vanilla merge_clin.R ex-data-9.adat clin-data.csv SampleId ex-data-9-merged.adat
```

```{r sys-call1, include = FALSE}
withr::with_dir(analysis_dir,
  base::system2(
    "Rscript",
    c("--vanilla",
      "merge_clin.R",
      "ex-data-9.adat",
      "clin-data.csv",
      "SampleId",
      "ex-data-9-merged.adat")
  )
)
```

```{r ls2}
dir(analysis_dir)
```


#### Alternative Syntax

In certain instances you may have the common key under a _different_ variable
name in their respective files. This is handled by a modification to
argument 3, which now takes the form `key1=key2` where `key1` contains the
common keys in the `*.adat` file, and `key2` contains keys for the `*.csv` file.

First let's create a new clinical data file with a different
variable name, `ClinID`:

```{r create-clin-2}
names(df) <- c("ClinID", "letter", "size")   # rename original `df`
df

# write clinical data to file
withr::with_dir(analysis_dir,
  write.csv(df, file = "clin-data2.csv", row.names = FALSE)
)
```

We can now execute the merge script at the command line as follows:

```bash
Rscript --vanilla merge_clin.R ex-data-9.adat clin-data2.csv SampleId=ClinID ex-data-9-merged2.adat
```

```{r sys-call2, include = FALSE}
withr::with_dir(analysis_dir,
  base::system2(
    "Rscript",
    c("--vanilla",
      "merge_clin.R",
      "ex-data-9.adat",
      "clin-data2.csv",
      "SampleId=ClinID",
      "ex-data-9-merged2.adat")
  )
)
```

```{r ls3}
dir(analysis_dir)
```

## Check Results

Now let's check that the merge was successful and yields the expected
`*.adat`:

```{r new-adat}
new <- withr::with_dir(analysis_dir,
  read_adat("ex-data-9-merged2.adat")
)
new

getMeta(new)

getAnalytes(new)
```


## Closing

Merging newly obtained clinical variables into existing 'SomaScan' ADATs
is easy via the `merge_clin.R` script provided with
[SomaDataIO](https://CRAN.R-project.org/package=SomaDataIO).
If you run into any trouble please do not hesitate to reach out
to <techsupport@somalogic.com> or
[file an issue](https://github.com/SomaLogic/SomaDataIO/issues/new) on
our [GitHub](https://github.com/SomaLogic/SomaDataIO) repository.


```{r teardown, include = FALSE}
if ( dir.exists(analysis_dir) ) {
  unlink(analysis_dir, force = TRUE)
}
```

---------------------


Created by [Rmarkdown](https://github.com/rstudio/rmarkdown)
(v`r utils::packageVersion("rmarkdown")`) and `r R.version$version.string`.
