---
title: "Getting Started"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(SomaDataIO)
library(purrr)
library(tibble)
library(ggplot2)
library(stringr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(101)
```


-------------------------

# SomaDataIO

The `SomaDataIO` package loads the SomaLogic, Inc. proprietary 
data file, called an ADAT file ("\*.adat"). It also provides auxiliary 
functions for extracting and manipulating relevant information from an 
ADAT, as well as exporting a modified ADAT to file.


-------------------------

## Main Features

* Loading data
    + Import a text file in the `*.adat` format into an `R` session
      as a `soma_adat` object.
* Manipulating data
    + Subset, reorder, and list various fields of a `soma_adat` object.
* Exporting data
    + Write out a `soma_adat` object as a `*.adat` text file.


----------------------


## Load an ADAT
```{r adat}
f <- system.file("sample", "sample.adat",
                 package = "SomaDataIO", mustWork = TRUE)
my_adat <- read_adat(f)
is.soma_adat(my_adat)

my_adat                      # print

print(my_adat, show_header = TRUE)  # if simply wish to see Header info, no features
```


## Attributes contain feature information
```{r atts}
attributes(my_adat)$Col.Meta
```


## Manipulate a `soma_adat` Object
```{r features}
getFeatures(my_adat) %>% head(20)     # first 20 features
getFeatures(my_adat) %>% length()     # how many features
getFeatures(my_adat, n = TRUE)        # the `n` argument; no. features
```


## Write out a `soma_adat`
```{r write}
is.intact.attributes(my_adat)         # attributes MUST be intact to write

fout <- tempfile(fileext = ".adat")   # random file name
write_adat(my_adat, file = fout)
```


# Typical Analysis
## Load data
```{r load}
data <- read_adat(f)
dim(data)

train <- dplyr::filter(data, TimePoint == "Young")
table(train$SampleGroup)    # M vs F
```

## Subset Features starting with `MMP`
```{r mmps}
mmps <- my_adat %>% 
  getFeatureData() %>% 
  filter(grepl("^MMP", Target))

# feature data info
mmps  
```

## Generate list of `t-tests`
```{r t_tests}
t_tests <- mmps$AptName %>% 
  purrr::set_names(., .) %>% 
  purrr::map(~ {
    sprintf("%s ~ SampleGroup", .x) %>% 
      as.formula() %>% 
      t.test(data = data)
  })
```

## Re-order by `p-value`
```{r p_values}
idx <- t_tests %>% 
  purrr::map("p.value") %>% 
  stringr::str_order()
idx

t_tests %<>% magrittr::extract(idx)
t_tests %>% head() %>% purrr::map("p.value")
```

## Visualize with `ggplot2()`
```{r ggplot_boxes, fig.width = 9, fig.height = 8}
targets <- as.list(mmps$AptName) %>%
  purrr::set_names(mmps$TargetFullName)
df <- data %>% 
  dplyr::select(SampleGroup, all_of(mmps$AptName)) %>% 
  dplyr::rename(!!! targets) %>% 
  tidyr::pivot_longer(-SampleGroup, names_to = "Feature", values_to = "RFU") %>% 
  tibble::as_tibble()
df

df %>% 
  ggplot(aes(x = SampleGroup, y = log10(RFU), colour = SampleGroup)) +
  geom_boxplot(alpha = 0.5) +
  facet_wrap(~ Feature)
```



-----------------------------


## MIT LICENSE

SomaDataIO™

Copyright © 2020 SomaLogic, Inc.

Permission is hereby granted, free of charge, to any person obtaining
a copy of the SomaDataIO software and associated documentation files
(the "Software"), to deal in the Software without restriction,
including without limitation the rights to use, copy, modify, merge,
publish, distribute, sub-license, and/or sell copies of the Software,
and to permit persons to whom the Software is furnished to do so,
subject to the following conditions outlined below.  Further,
SomaDataIO and SomaLogic are trademarks owned by SomaLogic, Inc.  No
license is hereby granted to these trademarks other than for purposes
of identifying the origin or source of the Software.

The above copyright notice and this permission notice shall be
included in all copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NON-INFRINGEMENT.
IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDER(S) BE LIABLE FOR
ANY CLAIM, DAMAGES, WHETHER DIRECT OR INDIRECT, OR OTHER LIABILITY,
WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS
IN THE SOFTWARE.


-----------------------------

Created by [Rmarkdown](https://github.com/rstudio/rmarkdown)
(v`r utils::packageVersion("rmarkdown")`) and `r R.version$version.string`.

