---
title: "Lifting and Bridging SomaScan"
author: "Stu Field, Standard BioTools, Inc."
description: >
  A primer on lifting and bridging 'SomaScan' data.
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Lifting and Bridging SomaScan}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
library(SomaDataIO)
library(ggplot2)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/lifting-"
)
calc_ccc <- function(x, y) {
  k   <- length(x)
  sdx <- sd(x)
  sdy <- sd(y)
  rho <- stats::cor(x, y, method = "pearson")
  v   <- sdx / sdy   # scale shift
  sx2 <- stats::var(x) * (k - 1) / k
  sy2 <- stats::var(y) * (k - 1) / k
  # location shift relative to scale
  u   <- ( mean(x) - mean(y) ) / ( (sx2 * sy2)^0.25 )
  rho * ( (v + 1 / v + u^2 ) / 2 )^-1
}
```


# Overview

`SomaDataIO` contains functionality to bridge (aka "lift") between
various SomaScan versions by linear transformations of RFU data.
Lifting between various versions is essentially a calibration of the
analytes/features in RFU space.


## Why lift?

The SomaScan platform continually improves its technical processes
between assay versions. The primary change of interest is content expansion,
and other protocol changes may be implemented including: changing reagents,
liquid handling equipment, and well volumes.

For any given analyte, these technical upgrades may result
in minute measurement signal differences, requiring a
calibration (aka "lifting" or "bridging") to bring RFU
values into a comparable signal space. This is accomplished
by applying an analyte-specific scalar, a linear transformation,
to each analyte RFU measurement (column).

### Current SomaScan Versions

| **Version**   | **Commercial Name** | **Size**      |
|:------------- |:-------------------:| -------------:|
|  `V4`         | 5k                  | 5284          |
|  `v4.1`       | 7k                  | 7596          |
|  `v5.0`       | 11k                 | 11083         |


### Lifting Requirements

There are 4 main requirements in order to reliably bridge
across SomaScan signal space:

1. the `soma_adat` object attributes, where SomaScan signal information is
   stored, must be intact (see `is_intact_attr()`).
1. the sample matrix must be either human serum or human EDTA-plasma.
   No other matrices are currently supported. Additionally, bridging
   must *not* be applied across matrices (i.e. serum $\leftrightarrow$ plasma).
1. the RFU data must have been normalized by Adaptive Normalization via
   Maximum-Likelihood (ANML). This is the standard normalization for
   most SomaScan deliveries.
1. the current SomaScan version and signal space must be one of those
   above (see table), i.e. one of `5k`, `7k`, or `11k`. Older versions
   of SomaScan are not supported.

---------------

## Lifting Scalars

Lifting (aka "bridging") scalars are numeric values used to multiply a
vector of RFU values to linearly transform them into another signal space.

Lifting scalars are generated from matched samples (n $>$ 1000) from a
healthy, normal reference population were run across assay versions.
This experiment was run separately for both serum and plasma and all
SomaScan runs were first normalized as per the standard normalization
procedure, and flagged samples were removed prior to further analysis.

For each analyte, the lifting scalar is computed as the ratio of
population _medians_ between assay versions. For example,
the linear scalar for the $i^{th}$ analyte translating from
`11k` $\rightarrow$ `7k` is defined as:

$$
R_i = \frac{\hat\mu_{7k}}{\hat\mu_{11k}},
$$

where $\hat\mu$ is the _median_ signal for the $i^{th}$ analyte.
Signals generated in `11k` space can be multiplied by this scale factor
to translate into `7k` space.

Below is a concordance plot of what this shift would look like for a single
analyte on a _simulated_ reference population. Please see the section below
on Lin's CCC for its definition and interpretation.

```{r lift-concord, echo = FALSE, fig.width = 6, fig.height = 4, fig.cap = "Figure 1. Signal concordance for a single analyte pre- and post-lifting."}
rfu  <- dplyr::filter(example_data, SampleType == "Sample")$seq.9016.12
L    <- length(rfu)
rfu2 <- rfu +
  withr::with_seed(123, rnorm(L, mean = 500, sd = sd(rfu) / 3))
sf   <- median(rfu) / median(rfu2)
pre  <- data.frame(x = rfu, y = rfu2)
pre$group <- sprintf("pre-lift (%0.3f)", calc_ccc(pre$x, pre$y))
post <- data.frame(x = rfu, y = rfu2 * sf)
post$group <- sprintf("post-lift (%0.3f)", calc_ccc(post$x, post$y))
plot_df <- rbind(pre, post)
plot_df$group <- factor(plot_df$group, levels = rev(sort(unique(plot_df$group))))
lims <- range(plot_df[, -3L])
plot_df |>
  ggplot(aes(x = x, y = y, colour = group)) +
  geom_point(alpha = 0.5, size = 3) +
  scale_x_log10(guide = "axis_logticks") +
  scale_y_log10(guide = "axis_logticks") +
  scale_colour_manual(name = "CCC", values = c("#00A499", "#24135F")) +
  expand_limits(x = lims, y = lims) +
  labs(x = "SomaScan 7k", y = "SomaScan 11k",
       title = sprintf("Lifting Concordance (Scalar = %0.3f)", sf)) +
  geom_abline(slope = 1, intercept = 0, color = "black")
```

---------------

## Lifting Concordance

Measurements generated from the matched samples used to calculate
the lifting scalars were also used to calculate the post-hoc
Lin's Concordance Correlation Coefficient (CCC) estimates
of the SomaScan bridge.

Lin's CCC is calculated by computing the correlation between
post-lift RFU values and the RFU values generated on the
original SomaScan version, and is defined by:

$$
CCC = \frac{2\rho\hat\sigma_x\hat\sigma_y}{(\hat\mu_x - \hat\mu_y)^2 + \hat\sigma^2_x + \hat\sigma^2_y},
$$

where $\rho$, $\hat\mu$, and $\hat\sigma$ are the Pearson correlation
coefficient, and the estimated mean and standard deviation from
assay version groups _x_ and _y_ respectively.


### Interpretation of CCC

Lin's CCC was chosen to evaluate lifting performance because it is 
characterized not only by correlation (Pearson's $\rho$), but
also accounts for deviation from the $y = x$ unit line (diagonal). 
CCC range is in $[-1, 1]$ and can be viewed as an estimate of the
confidence in the bridging transformation (in normal reference
samples) across SomaScan versions.
Examples of factors that could affect lifting CCC are:

- analytes/reagents with high intra-assay CV (Coefficient of Variation)
- analytes/reagents signaling near background or saturation levels


### Accessing CCC

The `getSomaScanLiftCCC()` function retrieves these values
from an internal object for either `"serum"` and `"plasma"`.

```{r ccc}
plasma <- getSomaScanLiftCCC("p")
plasma

serum <- getSomaScanLiftCCC("s")
serum
```

```{r cdf-ccc, fig.width = 6, fig.height = 5, fig.cap = "Figure 2. Cumulative distribution function of CCC values for the 11k -> 7k lift."}
cdf_df <- data.frame(
  ccc    = c(plasma$plasma_11k_to_7k_ccc, serum$serum_11k_to_7k_ccc),
  matrix = rep(c("plasma", "serum"), each = nrow(plasma))
)
cdf_df <- cdf_df[!is.na(cdf_df$ccc), ]   # rm NAs; non-comparable analytes
ggplot(cdf_df, aes(x = ccc, colour = matrix)) +
  stat_ecdf(linewidth = 0.75) +
  scale_colour_manual(name = "", values = c("#00A499", "#24135F")) +
  labs(title = "CDF of CCC Values",
       x = "Lin's CCC (11k -> 7k)", y = "P(X < x)") +
  coord_cartesian()
```

As shown in distribution above, for the `11k` $\rightarrow$ `7k` lift,
post-bridging CCC values above 0.75 (considered high quality) are
approximately 88% and 84% of the SomaScan menu for plasma
and serum respectively. In fact, characterizing CCC lifting quality into 3
categories (Low, Medium, High) yields the table below:

```{r ecdf-table, echo = FALSE}
fn <- function(x) {
  cdf <- stats::ecdf(x)
  data.frame(lo  = cdf(0.5), med = cdf(0.75) - cdf(0.5), hi  = 1 - cdf(0.75))
}
do.call(rbind, tapply(cdf_df$ccc, cdf_df$matrix, fn)) |>
  round(3L) |>
  set_rn(c("Plasma", "Serum")) |>
  rn2col("Matrix") |>
  knitr::kable(
    col.names = c("Matrix", "Low [0, 0.5)", "Medium [0.5, 0.75)", "High [0.75, 1]"),
    caption = "Table 1. The proportion of the SomaScan menu split into 3 categories by CCC."
  )
```


## SomaScan Analyte Setdiff

For any given bridge, there is a common, intersecting subset of
analytes between SomaScan versions. Non-intersecting analytes will be
either missing or added in the new signal space. As a result,
bridging data across SomaScan may involve either skipping analytes (columns)
or scaling by 1.0. `SomaDataIO` has internal checks that trigger
warnings if these conditions are met.

There are two scenarios to consider:

* Newer versions of SomaScan typically have additional content,
  i.e. new reagents added to the multi-plex assay that bind to additional
  proteins. When lifting _to_ a previous SomaScan version,
  new reagents that do _not_ exist in the "earlier" assay version
  assay are scaled by 1.0, and thus are maintained, unmodified in
  the returned object.
  Downstream analysis may require removing these columns in order
  to combine these data with a previous study from an earlier
  SomaScan version, e.g. with `collapseAdats()`.
* In the inverse scenario, lifting "forward" _from_ a previous, lower-plex
  version, there will be extra reference values that are unnecessary
  to perform the lift, and a warning is triggered. The resulting data
  consists of RFU data in the "new" signal space, but with fewer analytes
  than would otherwise be expected (e.g. `11k` space with only 5284
  analytes; see example below).


------------


## Example: `5k` $\rightarrow$ `11k`

Since `example_data` object was originally run on SomaScan 
`r getSignalSpace(example_data)`, this vignette will demonstrate
the lifting/bridging process _from_ a `5k` $\rightarrow$ `11k`
signal space, the most recent SomaScan version.

### Steps

1. Determine that attributes are intact.
   ```r
   is_intact_attr(adat)
   ```
1. Determine the matrix type of the data (serum or plasma).
   ```r
   attr(adat, "Header.Meta")$HEADER$StudyMatrix
   ```
1. Ensure the current SomaScan signal space is lift-supported.
   ```r
   getSignalSpace(adat)
   checkSomaScanVersion(getSignalSpace(adat))
   ```
1. Apply analyte-specific scalars to their corresponding
   columns via `lift_adat()`.
   ```r
   lift_adat(adat, bridge = "<direction>")
   ```
   Current `bridge` options are: `r dQuote(eval(formals(lift_adat)$bridge))`.


### Step 1
```{r attr}
# determine intact attributes
# must be TRUE
is_intact_attr(example_data)
```

### Step 2
```{r matx}
# determine study matrix
# must be Human Serum or EDTA-Plasma
attr(example_data, "Header.Meta")$HEADER$StudyMatrix |> as.character()
```

Confirm that the matrix of the SomaScan run was `"EDTA Plasma"`:

### Step 3
```{r version}
# determine if current space can be lifted
# must be V4, v4.1, or v5.0
from_space <- getSignalSpace(example_data)
from_space

# must be NULL
is.null(checkSomaScanVersion(from_space))
```

Finally, invoke `lift_adat()` to perform the bridge/transformation:

### Step 4
```{r lift}
lift_11k <- lift_adat(example_data, bridge = "5k_to_11k")

is_lifted(lift_11k)           # signal space was lifted

is.soma_adat(lift_11k)        # preserves 'soma_adat' class

getSignalSpace(lift_11k)      # current space

getSomaScanVersion(lift_11k)  # original space
```


-------------


## Caveats to Consider

### Was the SomaScan bridge successful?

Lifting SomaScan involves a simple linear transformation of a
numeric vector (of RFU values), thus in one sense it will always
be "successful". However, users often wish to know if this was
the correct course of action for their data. 

From the concordance plot in **Figure 1**, we can see that the
transformation is *reducing* the `11k` RFU brightness
by ~`r round(100*(1 - sf))`% in accordance with the median signal difference
that existed in the reference population (of healthy normals).
Rare edge cases aside, this is *usually* the desired outcome, otherwise
downstream analysis would be confounded by the uncorrected shift in
SomaScan space, and would likely result in significant differences related
to signal space rather than actual biology.

### Should you filter analytes?

Users often ask if certain analytes be should removed based on a
given CCC threshold *prior* to analysis. 
The issue of choosing an appropriate threshold aside, unless
there is prior knowledge justifying removal, we do not recommend 
removing analytes based on CCC alone.

This advice stems from how the CCC values are initially calculated; i.e.
from a healthy, normal reference population sampled across two
versions of SomaScan. Recall that CCC is influenced by CV and thus
signaling range. For example, if a given analyte is near its
limit of detection in a healthy population, and therefore likely
has a high(er) CV, i.e. low CCC, removing this analyte may *not* be the
desired course of action in a disease population where that
analyte could be signaling in the linear range.

Therefore, we currently recommend careful evaluation on a
case by case basis using prior knowledge and orthogonal
justification before filtering analytes from discovery or
exploratory analyses.


-------------

## Questions

As always, if you have any bridging or lifting questions,
we are here to help. Please reach out to us via:

* via GitHub [SUPPORT](https://somalogic.github.io/SomaDataIO/SUPPORT.html)
* Global Scientific Engagement Team: <techsupport@somalogic.com>
* General SomaScan inquiries: <support@somalogic.com>
