---
title: "Binary Classification"
author: "Stu Field, SomaLogic Operating Co., Inc."
description: >
  Typical binary classification of 'SomaScan' data.
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Binary Classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/classify-"
)
```


----------------


## Classification via Logistic Regression


Although targeted statistical analyses are beyond the scope of 
the `SomaDataIO` package, below is an example analysis
that typical users/customers would perform on 'SomaScan' data.

It is not intended to be a definitive guide in statistical
analysis and existing packages do exist in the `R` ecosystem that perform
parts or extensions of these techniques. Many variations of the workflow
below exist, however the framework highlights how one could perform standard
_preliminary_ analyses on 'SomaScan' data.


## Load Libraries

```{r load-lib, message = FALSE}
library(SomaDataIO)
library(dplyr)
library(tidyr)
library(purrr)
```

## Data Preparation
```{r data-prep, fig.width = 7, fig.height = 3, fig.align = "center"}
# the `example_data` .adat object
# download from `SomaLogic-Data` repo or directly via bash command:
# `wget https://raw.githubusercontent.com/SomaLogic/SomaLogic-Data/main/example_data.adat`
# then read in to R with:
# example_data <- read_adat("example_data.adat")
dim(example_data)

table(example_data$SampleType)

# prepare data set for analysis using `preProcessAdat()`
cleanData <- example_data |>
  preProcessAdat(
    filter.features = TRUE,      # remove non-human protein features
    filter.controls = TRUE,      # remove control samples
    filter.rowcheck = TRUE,      # retain only passing samples
    log.10          = TRUE,      # log10 transform
    center.scale    = TRUE       # center/scale analytes
  )

# drop any missing values in Sex, and convert to binary 0/1 variable
cleanData <- cleanData |> 
  drop_na(Sex) |>                              # rm NAs if present
  mutate(Group = as.numeric(factor(Sex)) - 1)  # map Sex -> 0/1

table(cleanData$Sex)

table(cleanData$Group)    # F = 0; M = 1
```

## Set up Train/Test Data

```{r train-test}
# idx = hold-out 
# seed resulting in 50/50 class balance
idx   <- withr::with_seed(3, sample(1:nrow(cleanData), size = nrow(cleanData) - 50))
train <- cleanData[idx, ]
test  <- cleanData[-idx, ]

# assert no overlap
isTRUE(
  all.equal(intersect(rownames(train), rownames(test)), character(0))
)
```


## Logistic Regression
We use the `cleanData`, `train`, and `test` data objects from above.

### Predict Sex
```{r logreg-tbl}
LR_tbl <- getAnalyteInfo(train) |>
  select(AptName, SeqId, Target = TargetFullName, EntrezGeneSymbol, UniProt) |>
  mutate(
    formula  = map(AptName, ~ as.formula(paste("Group ~", .x))),  # create formula
    model    = map(formula, ~ stats::glm(.x, data = train, family = "binomial", model = FALSE)),  # fit glm()
    beta_hat = map(model, coef) |> map_dbl(2L),     # pull out coef Beta
    p.value  = map2_dbl(model, AptName, ~ {
      summary(.x)$coefficients[.y, "Pr(>|z|)"] }),  # pull out p-values
    fdr      = p.adjust(p.value, method = "BH")     # FDR correction multiple testing
  ) |>
  arrange(p.value) |>            # re-order by `p-value`
  mutate(rank = row_number())    # add numeric ranks

LR_tbl
```


### Fit Model | Calculate Performance

Next, select features for the model fit. We have a good idea of reasonable `Sex`
markers from prior knowledge (`CGA*`), and fortunately many of these are highly
ranked in `LR_tbl`. Below we fit a 4-marker logistic regression model from
cherry-picked gender-related features:

```{r fit-logreg}
# AptName is index key between `LR_tbl` and `train`
feats <- LR_tbl$AptName[c(1L, 3L, 5L, 7L)]
form  <- as.formula(paste("Group ~", paste(feats, collapse = "+")))
fit   <- glm(form, data = train, family = "binomial", model = FALSE)
pred  <- tibble(
  true_class = test$Sex,                                         # orig class label
  pred       = predict(fit, newdata = test, type = "response"),  # prob. 'Male'
  pred_class = ifelse(pred < 0.5, "F", "M"),                     # class label
)
conf <- table(pred$true_class, pred$pred_class, dnn = list("Actual", "Predicted"))
tp   <- conf[2L, 2L]
tn   <- conf[1L, 1L]
fp   <- conf[1L, 2L]
fn   <- conf[2L, 1L]

# Confusion matrix
conf

# Classification metrics
tibble(Sensitivity = tp / (tp + fn),
       Specificity = tn / (tn + fp),
       Accuracy    = (tp + tn) / sum(conf),
       PPV         = tp / (tp + fp),
       NPV         = tn / (tn + fn)
)
```

