---
title: "Two-Group Comparison"
author: "Stu Field, Standard BioTools, Inc."
description: >
  Typical 2-group comparison of 'SomaScan' data.
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Two-Group Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(SomaDataIO)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/two-group-"
)
```


--------------


## Differential Expression via *t*-test

Although targeted statistical analyses are beyond the scope of 
the `SomaDataIO` package, below is an example analysis
that typical users/customers would perform on 'SomaScan' data.

It is not intended to be a definitive guide in statistical
analysis and existing packages do exist in the `R` ecosystem that perform
parts or extensions of these techniques. Many variations of the workflow
below exist, however the framework highlights how one could perform standard
_preliminary_ analyses on 'SomaScan' data.


## Data Preparation
```{r data-prep, fig.width = 7, fig.height = 3, fig.align = "center"}
# the `example_data` .adat object
# download from `SomaLogic-Data` repo or directly via bash command:
# `wget https://raw.githubusercontent.com/SomaLogic/SomaLogic-Data/main/example_data.adat`
# then read in to R with:
# example_data <- read_adat("example_data.adat")
dim(example_data)

table(example_data$SampleType)

# prepare data set for analysis using `preProcessAdat()`
cleanData <- example_data |>
  preProcessAdat(
    filter.features = TRUE,            # rm non-human protein features
    filter.controls = TRUE,            # rm control samples
    filter.qc       = TRUE,            # rm non-passing qc samples
    log.10          = TRUE,            # log10 transform
    center.scale    = TRUE             # center/scale analytes
  )

# drop any missing values in Sex
cleanData <- cleanData |>
  drop_na(Sex)                         # rm NAs if present

table(cleanData$Sex)
```



## Compare Two Groups (`M`/`F`)
### Get annotations via `getAnalyteInfo()`:

```{r get-anno}
t_tests <- getAnalyteInfo(cleanData) |>
  select(AptName, SeqId, Target = TargetFullName, EntrezGeneSymbol, UniProt)

# Feature data info:
#   Subset via dplyr::filter(t_tests, ...) here to
#   restrict analysis to only certain analytes

t_tests
```



### Calculate `t-tests`
Use a "list columns" approach via nested tibble object
using `dplyr`, `purrr`, and `stats::t.test()`

```{r t-tests}
t_tests <- t_tests |>
  mutate(
    formula = map(AptName, ~ as.formula(paste(.x, "~ Sex"))), # create formula
    t_test  = map(formula, ~ stats::t.test(.x, data = cleanData)),  # fit t-tests
    t_stat  = map_dbl(t_test, "statistic"),            # pull out t-statistic
    p.value = map_dbl(t_test, "p.value"),              # pull out p-values
    fdr     = p.adjust(p.value, method = "BH")         # FDR for multiple testing
  ) |>
  arrange(p.value) |>            # re-order by `p-value`
  mutate(rank = row_number())    # add numeric ranks

# View analysis tibble
t_tests
```



### Visualize with `ggplot2()`
Create a plotting tibble in the "long" format for `ggplot2`:

```{r ggplot-data}
target_map <- head(t_tests, 12L) |>     # mapping table
  select(AptName, Target)               # SeqId -> Target

plot_tbl <- example_data |>             # plot non-center/scale data
  filter(SampleType == "Sample") |>     # rm control samples
  drop_na(Sex) |>                       # rm NAs if present
  log10() |>                            # log10-transform for plotting
  select(Sex, target_map$AptName) |>    # top 12 analytes
  pivot_longer(cols = -Sex, names_to = "AptName", values_to = "RFU") |>
  left_join(target_map, by = "AptName") |>
  # order factor levels by 't_tests' rank to order plots below
  mutate(Target = factor(Target, levels = target_map$Target))

plot_tbl
```

```{r seed, include = FALSE}
# seed for geom::jitter() so Git isn't triggered every time
set.seed(1)
```

```{r ggplot-boxes, fig.width = 7, fig.height = 7, fig.align = "center"}
plot_tbl |>
  ggplot(aes(x = Sex, y = RFU, fill = Sex)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  scale_fill_manual(values = c("#24135F", "#00A499")) +
  geom_jitter(shape = 16, width = 0.1, alpha = 0.5) +
  facet_wrap(~ Target, ncol = 3) +
  ggtitle("Boxplots of Top Analytes by t-test") +
  labs(y = "log10(RFU)") +
  theme(plot.title = element_text(size = 21, face = "bold"),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.position = "top"
  )
```

