---
title: "Linear Regression"
author: "Stu Field, SomaLogic Operating Co., Inc."
description: >
  Typical linear regression of continuous 'SomaScan' data.
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Linear Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(SomaDataIO)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/linear-reg-"
)
```


--------------


## Regression of Continuous Variables

Although targeted statistical analyses are beyond the scope of 
the `SomaDataIO` package, below is an example analysis
that typical users/customers would perform on 'SomaScan' data.

It is not intended to be a definitive guide in statistical
analysis and existing packages do exist in the `R` ecosystem that perform
parts or extensions of these techniques. Many variations of the workflow
below exist, however the framework highlights how one could perform standard
_preliminary_ analyses on 'SomaScan' data.


## Data Preparation
```{r data-prep, fig.width = 7, fig.height = 3, fig.align = "center"}
# the `example_data` .adat object
# download from `SomaLogic-Data` repo or directly via bash command:
# `wget https://raw.githubusercontent.com/SomaLogic/SomaLogic-Data/main/example_data.adat`
# then read in to R with:
# example_data <- read_adat("example_data.adat")
dim(example_data)

table(example_data$SampleType)

# prepare data set for analysis using `preProcessAdat()`
cleanData <- example_data |>
  preProcessAdat(
    filter.features = TRUE,            # rm non-human protein features
    filter.controls = TRUE,            # rm control samples
    filter.qc       = TRUE,            # rm non-passing qc samples
    log.10          = TRUE,            # log10 transform
    center.scale    = TRUE             # center/scale analytes
  )

# drop any missing Age values
cleanData <- cleanData |>
  drop_na(Age)                         # rm NAs if present

summary(cleanData$Age)
```


## Set up Train/Test Data

```{r train-test}
# idx = hold-out 
idx   <- withr::with_seed(3, sample(1:nrow(cleanData), size = nrow(cleanData) - 50))
train <- cleanData[idx, ]
test  <- cleanData[-idx, ]

# assert no overlap
isTRUE(
  all.equal(intersect(rownames(train), rownames(test)), character(0))
)
```


## Linear Regression
We use the `cleanData`, `train`, and `test` data objects from above.

### Predict Age
```{r linreg-tbl}
LinR_tbl <- getAnalyteInfo(train) |>                # `train` from above
  select(AptName, SeqId, Target = TargetFullName, EntrezGeneSymbol, UniProt) |>
  mutate(
    formula = map(AptName, ~ as.formula(paste("Age ~", .x, collapse = " + "))),
    model   = map(formula, ~ stats::lm(.x, data = train, model = FALSE)), # fit models
    slope   = map(model, coef) |> map_dbl(2L),     # pull out B_1
    p.value = map2_dbl(model, AptName, ~ {
      summary(.x)$coefficients[.y, "Pr(>|t|)"] }), # pull out p-values
    fdr     = p.adjust(p.value, method = "BH")     # FDR for multiple testing
  ) |>
  arrange(p.value) |>            # re-order by `p-value`
  mutate(rank = row_number())    # add numeric ranks

LinR_tbl
```


### Fit Model | Calculate Performance
Fit an 8-marker model with the top 8 features from `LinR_tbl`:

```{r linreg-fit}
feats <- head(LinR_tbl$AptName, 8L)
form  <- as.formula(paste("Age ~", paste(feats, collapse = "+")))
fit   <- stats::lm(form, data = train, model = FALSE)
n     <- nrow(test)
p     <- length(feats)

# Results
res   <- tibble(
  true_age   = test$Age,
  pred_age   = predict(fit, newdata = test),
  pred_error = pred_age - true_age
)

# Lin's Concordance Correl. Coef.
# Accounts for location + scale shifts
linCCC <- function(x, y) {
  stopifnot(length(x) == length(y))
  a <- 2 * cor(x, y) * sd(x) * sd(y)
  b <- var(x) + var(y) + (mean(x) - mean(y))^2
  a / b
}

# Regression metrics
tibble(
  rss  = sum(res$pred_error^2),                 # residual sum of squares
  tss  = sum((test$Age - mean(test$Age))^2),    # total sum of squares
  rsq  = 1 - (rss / tss),                       # R-squared
  rsqadj = max(0, 1 - (1 - rsq) * (n - 1) / (n - p - 1)), # Adjusted R-squared
  R2   = stats::cor(res$true_age, res$pred_age)^2,        # R-squared Pearson approx.
  MAE  = mean(abs(res$pred_error)),             # Mean Absolute Error
  RMSE = sqrt(mean(res$pred_error^2)),          # Root Mean Squared Error
  CCC  = linCCC(res$true_age, res$pred_age)     # Lin's CCC
)
```


### Visualize Concordance
```{r linreg-plot, fig.width = 7, fig.height = 7, fig.align = "center"}
lims <- range(res$true_age, res$pred_age)
res |>
  ggplot(aes(x = true_age, y = pred_age)) +
  geom_point(colour = "#24135F", alpha = 0.5, size = 4) +
  expand_limits(x = lims, y = lims) +                # make square
  geom_abline(slope = 1, colour = "black") +         # add unit line
  geom_rug(colour = "#286d9b", linewidth = 0.2) +
  labs(y = "Predicted Age", x = "Actual Age") +
  ggtitle("Concordance in Predicted vs. Actual Age") +
  theme(plot.title = element_text(size = 21, face = "bold"),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14))
```


