---
title: "ANOVA Three-Group Analysis"
author: "Stu Field, Standard BioTools, Inc."
description: >
  Typical > 2-group comparison of 'SomaScan' data.
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{ANOVA Three-Group Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
library(SomaDataIO)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/three-group-"
)
```


--------------


## Differential Expression via ANOVA

Although targeted statistical analyses are beyond the scope of 
the `SomaDataIO` package, below is an example analysis
that typical users/customers would perform on 'SomaScan' data.

It is not intended to be a definitive guide in statistical
analysis and existing packages do exist in the `R` ecosystem that perform
parts or extensions of these techniques. Many variations of the workflow
below exist, however the framework highlights how one could perform standard
_preliminary_ analyses on 'SomaScan' data.


## Data Preparation
```{r data-prep}
# the `example_data` package data
dim(example_data)

table(example_data$SampleType)

# center/scale
cs <- function(.x) {    # .x = numeric vector
  out <- .x - mean(.x)  # center
  out / sd(out)         # scale
}

# prepare data set for analysis
cleanData <- example_data |>
  filter(SampleType == "Sample") |>         # rm control samples
  drop_na(Sex) |>                           # rm NAs if present
  log10() |>                                # log10-transform (Math Generic)
  modify_at(getAnalytes(example_data), cs)  # center/scale analytes

# dummy 3 group setup
# set up semi-random 3-group with structure
# based on the `Sex` variable (with known structure)
cleanData$Group <- ifelse(cleanData$Sex == "F", "A", "B")
g3 <- withr::with_seed(123, sample(1:nrow(cleanData), size = round(nrow(cleanData) / 3)))
cleanData$Group[g3] <- "C"
table(cleanData$Group)
```



## Compare Three Groups (`A`/`B`/`C`)
### Get annotations via `getAnalyteInfo()`:

```{r get-anno}
aov_tbl <- getAnalyteInfo(cleanData) |>
  select(AptName, SeqId, Target = TargetFullName, EntrezGeneSymbol, UniProt)

# Feature data info:
#   Subset via dplyr::filter(aov_tbl, ...) here to
#   restrict analysis to only certain analytes

aov_tbl
```



### Calculate ANOVAs
Use a "list columns" approach via nested tibble object
using `dplyr`, `purrr`, and `stats::aov()`

```{r anova-models}
aov_tbl <- aov_tbl |>
  mutate(
    formula   = map(AptName, ~ as.formula(paste(.x, "~ Group"))), # create formula
    aov_model = map(formula, ~ stats::aov(.x, data = cleanData)),  # fit ANOVA-models
    aov_smry  = map(aov_model, summary) |> map(1L),      # summary() method
    F.stat    = map(aov_smry, "F value") |> map_dbl(1L), # pull out F-statistic
    p.value   = map(aov_smry, "Pr(>F)") |> map_dbl(1L),  # pull out p-values
    fdr       = p.adjust(p.value, method = "BH")         # FDR multiple testing
  ) |>
  arrange(p.value) |>            # re-order by `p-value`
  mutate(rank = row_number())    # add numeric ranks

# View analysis tibble
aov_tbl
```



### Visualize with `ggplot2()`
Create a plotting tibble in the "long" format for `ggplot2`:

```{r ggplot-data}
target_map <- head(aov_tbl, 12L) |>     # mapping table
  select(AptName, Target)               # SeqId -> Target

plot_tbl <- cleanData |>
  select(Group, target_map$AptName) |>    # top 12 analytes
  pivot_longer(cols = -Group, names_to = "AptName", values_to = "RFU") |>
  left_join(target_map, by = "AptName") |>
  # order factor levels by 'aov_tbl' rank to order plots below
  mutate(Target = factor(Target, levels = target_map$Target))

plot_tbl
```

```{r ggplot-pdfs, fig.width = 7, fig.height = 7, fig.align = "center"}
plot_tbl |>
  ggplot(aes(x = RFU, fill = Group)) +
  geom_density(linetype = 0, alpha = 0.25) +
  scale_fill_manual(values = c("#24135F", "#00A499", "#006BA6")) +
  facet_wrap(~ Target, ncol = 3) +
  ggtitle("Probability Density of Top Analytes by ANOVA") +
  labs(y = "log10(RFU)") +
  theme(plot.title = element_text(size = 21, face = "bold"),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.position = "top"
  )
```


