---
title: "Loading and Wrangling 'SomaScan'"
author: "Stu Field, Standard BioTools, Inc."
description: >
  How to load and manipulate a 'SomaScan' flat text file into
  and R environment.
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Loading and Wrangling 'SomaScan'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE, results = FALSE, message = FALSE}
options(width = 80)
#Sys.setlocale("LC_COLLATE", "C")
Sys.setlocale("LC_COLLATE", "en_US.UTF-8") # ensure common sorting envir
library(SomaDataIO)
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/wrangling-"
)
```


## Loading an ADAT

Load an ADAT text file into `R` memory with:

```{r read-adat}
# path to *.adat file
# replace with your file path
adat_path <- system.file("extdata", "example_data10.adat",
                         package = "SomaDataIO", mustWork = TRUE)
adat_path

my_adat <- read_adat(adat_path)

# class test
is.soma_adat(my_adat)

# S3 print method forwards -> tibble
my_adat

print(my_adat, show_header = TRUE)  # if simply wish to see Header info

# S3 summary method
# View Target and summary statistics
seqs <- tail(names(my_adat), 3L)
summary(my_adat[, seqs])

# Summarize by Sex
my_adat[, seqs] |>
  split(my_adat$Sex) |>
  lapply(summary)
```

### Debugging

Occasionally "problematic" ADATs can be difficult to parse. For this
purpose a convenient `debug = TRUE` argument to `read_adat()` allows you
to inspect the file specifications that `R` _thinks_ exist in the file.
This can be useful in identifying where/why/how a parse failure has occurred.
It is recommended to view this output and compare to the physical
text file itself to identify any misidentified or mismatched landmarks:

```{r debug}
read_adat(adat_path, debug = TRUE)
```


---------------------


## Wrangling
### Attributes Contain File and Feature Information

```{r atts}
names(attributes(my_adat))

# The `Col.Meta` attribute contains
# target annotation information
attr(my_adat, "Col.Meta")
```

### Analyte Features (`seq.xxxx.xx`)

```{r feats}
getAnalytes(my_adat) |> head(20L)    # first 20 analytes; see AptName above
getAnalytes(my_adat) |> length()     # how many analytes
getAnalytes(my_adat, n = TRUE)       # the `n` argument; no. analytes
```

### Feature Data

The `getAnalyteInfo()` function creates a lookup table that links
analyte feature names in the `soma_adat` object to the annotation
data in `?Col.Meta` via the common index-key, `AptName`, in column 1:

```{r annotations}
getAnalyteInfo(my_adat)
```


### Clinical Data

```{r meta}
getMeta(my_adat)             # clinical meta data for each sample
getMeta(my_adat, n = TRUE)   # also an `n` argument
```


### ADAT structure

The `soma_adat` object also contains specific structure that are useful
to users. Please also see `?colmeta` or `?annotations` for further
details about these fields.

---------------------



### Group Generics

You may perform basic mathematical transformations on the feature data _only_
with special `soma_adat` S3 methods (see `?groupGenerics`):

```{r group-generics}
head(my_adat$seq.2429.27)

logData <- log10(my_adat)    # a typical log10() transform
head(logData$seq.2429.27)

roundData <- round(my_adat)
head(roundData$seq.2429.27)

sqData <- sqrt(my_adat)
head(sqData$seq.2429.27)

antilog(1:4)

sum(my_adat < 100)  # low signalling values

all.equal(my_adat, sqrt(my_adat^2))

all.equal(my_adat, antilog(log10(my_adat)))
```


#### Math Generics

```{r math}
getGroupMembers("Math")

getGroupMembers("Compare")

getGroupMembers("Arith")

getGroupMembers("Summary")
```


### Full Complement of [dplyr](https://dplyr.tidyverse.org) S3 Methods

The `soma_adat` also comes with numerous class specific methods to the most
popular [dplyr](https://dplyr.tidyverse.org) generics that make working
with `soma_adat` objects simpler for those familiar with this standard toolkit:

```{r dplyr}
dim(my_adat)
males <- dplyr::filter(my_adat, Sex == "M")
dim(males)

males |>
  dplyr::select(SampleType, SampleMatrix, starts_with("NormScale"))
```


#### Merging Sample Annotation Data

The `example_data` object includes some sample annotation data built-in, with
the variables `Age` and `Sex` included for clinical samples, but in practice 
ADAT files generally do not have any clinical or sample annotation data fields
included.

To merge sample annotation data into an existing `soma_adat` class object,
use the `left_join()` method.  Here, joining the `ex_clin_data` `tibble` object
adds in two additional clinical variables, `smoking_status` and `alcohol_use`:

```{r merge-annotations}
# `clin_path` should be the elaborated path and file name of the *.csv or
# similar file to be loaded into the R workspace from your local file system
# (e.g. clin_path = "PATH_TO_CLIN/clin_data.csv")
# clin_data <- readr::read_csv(clin_path)

merged_adat <- my_adat |> 
  dplyr::filter(SampleType == "Sample") |> 
  dplyr::left_join(ex_clin_data, by = "SampleId") 

merged_adat |> 
  dplyr::select(SampleId, Age, Sex, smoking_status, alcohol_use) |> 
  head(n = 3)
```


### Available S3 Methods `soma_adat`

```{r methods}
# see full complement of `soma_adat` methods
methods(class = "soma_adat")
```


---------------------

## Writing a `soma_adat`

```{r write}
is_intact_attr(my_adat)   # MUST have intact attrs

write_adat(my_adat, file = tempfile("my-adat-", fileext = ".adat"))
```


