---
title: "Common train-test data setups"
author: "Stu Field, SomaLogic Operating Co., Inc."
description: >
  Simple code syntax for common machine learning (ML)
  training vs test setups in R.
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Common train-test data setups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(dplyr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 90)
```


# Introduction

Most machine learning (ML) analyses require a random split of original data
into training/test data sets, where the model is fit on the training data
and performance is evaluated on the test data set. The split proportions can
vary, though 80/20 training/test is common. It often depends on the number of
available samples and the class distribution in the splits.

Among many alternatives, there are 3 common approaches, all are equally
viable and depend on the analyst's weighing of pros/cons of each.
I recommend one of these below:

1. base R data frame indexing with [sample()] and `[`
1. use `dplyr::slice_sample()` or `dplyr::sample_frac()` in
   combination with `dplyr::anti_join()`
1. use the [rsample](https://rsample.tidymodels.org) package (not demonstrated)

---------


## Original Raw Data

In most analyses, you typically start with a raw original data set that
you must split randomly into training and test sets.

```{r raw-data}
raw <- SomaDataIO::rn2col(head(mtcars, 10L), "CarModel") |>
  SomaDataIO::add_rowid("id") |> # set up identifier variable for the join()
  tibble::as_tibble()
raw
```


---------


## Option #1: `sample()`

```{r train-test1}
n     <- nrow(raw)
idx   <- withr::with_seed(1, sample(1:n, floor(n / 2))) # sample random 50% (n = 5)
train <- raw[idx, ]
test  <- raw[-idx, ]
train

test
```


## Option #2: `anti_join()`

```{r train-test2}
# sample random 50% (n = 5)
train <- withr::with_seed(1, dplyr::slice_sample(raw, n = floor(n / 2)))

# or using `dplyr::sample_frac()`
# train <- withr::with_seed(1, dplyr::sample_frac(raw, size = 0.5))

# use anti_join() to get the sample setdiff
test <- dplyr::anti_join(raw, train, by = "id")
train

test
```
